---
title: "Exploring North American Bigfoot Sightings"
listing:
  contents: posts
  feed: true
  sort: "date desc"
  type: default
  categories: true
  sort-ui: false
  filter-ui: false
page-layout: full
title-block-banner: true
---

```{r}
#| message: true
#| warning: true

print("Starting library load")

library(leaflet)
print("leaflet OK")

library(readr)
print("readr OK")

library(dplyr)
print("dplyr OK")

library(lubridate)
print("lubridate OK")

library(sf)
print("sf OK")

library(tigris)
print("tigris OK")

library(ggplot2)
print("ggplot2 OK")

library(broom)
print("broom OK")

library(ggeffects)
print("ggeffects OK")

library(plotly)
print("plotly OK")

print("All libraries loaded successfully")
```

## **Introduction**

*Do you believe in Bigfoot?* Possibly the most famous North American cryptid, Bigfoot sightings have become so ubiquitous that an entire organization of researchers was founded to investigate the credibility of these claims, and even venture into describing the ecology of their alleged creature. This organization is known as the BFRO (Bigfoot Field Researchers Organization).

Every reported Bigfoot sighting is stored in the BFRO's website, bfro.net. Here, BFRO's scholars record the witness's story on the encounter, collect information on variables related to climate and location, and classify the encounter into Class A, Class B, or Class C. Class A encounters are the most definitive, ruling out animals, lying/impaired witnesses, or otherwise "misidentifications". Class B encounters are less credible, with the potential for misidentifications or unclear evidence. And Class C encounters may be second or third-hand accounts, with little or untraceable evidence.

On September 13th, 2022, TidyTuesday provided a dataset scraped from the BFRO's catalogue of Bigfoot encounters to their weekly blog. This dataset contains many of the aforementioned climate and location variables as well as their credibility class and- *very* helpfully- the approximate latitude and longitude of the encounter.

Here I read in the data;

```{r}
#| message: false
#| warning: false

# install.packages("tidytuesdayR")

tuesdata <- tidytuesdayR::tt_load('2022-09-13')
tuesdata <- tidytuesdayR::tt_load(2022, week = 37)

bigfoot <- tuesdata$bigfoot

# Or read in the data manually

bigfoot <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2022/2022-09-13/bigfoot.csv')
```

Some of the variables provide a clear path towards a couple different visualizations answering a couple of different questions. My main goal with this project was answer the who, what, when, where, and whys of Bigfoot sightings over time. We have plenty of geographic variables, including longitudes/latitudes **and** the state and county where it occured. With these, I would like to generate a map that can be used to see each individual sighting's story and a density map displaying which states have the most sightings to answer the Whats, Whos, and Wheres. From there, we can examine how trends in sightings have changed over time, answering the Whens. And finally we can use some of those climate variables to get an idea of Why encoutners occur.

## **Data Tidying**

First I just want to get an idea of how many sightings we have, and where they occured, as well as just the narrative for the encounter. This visualization is more about exploring, not necessarily to support any kind of statistical conclusions.

Here I tidy the data for this leaflet map;

```{r}
bigfoot_stories <- bigfoot %>%   filter(!is.na(observed)) %>%   filter(!is.na(location_details))
```

\*note: 789 observations were removed at this stage for having missing observation text and location detail.

## **Visualizations**

Leaflet map:

```{r}
content <- bigfoot_stories %>%
  mutate(popup = paste0('<a href =', bigfoot_stories$location_details, '>',
                        bigfoot_stories$observed, '</a>'))
```

```{r}
bigfoot_map <- leaflet(bigfoot_stories) |>
  setView(lng = -98.583, lat = 39.833, zoom = 1) |>
  addTiles() |>
  addProviderTiles(providers$Esri.WorldPhysical) |>
  addMarkers(lng = bigfoot_stories$longitude, lat = bigfoot_stories$latitude,
             clusterOptions = markerClusterOptions(),
             popup = content$popup)
```

\*note: 976 sightings were removed from this interactive map as they did not provide the longitudes and latitudes.

```{r}
bigfoot_map
```

This map is nice because you can zoom in on each encounter and see *exactly* where each Bigfoot encounter occured. But because of the subgrouping of the pins, it is difficult to determine the geographic density of these sightings. It certainly *looks* like there may be some patterns. For example, in the Midwest, there appears to be a somewhat lower number of pins when you zoom out to look at the entire country, but it is difficult to determine for sure. We have the *stories*, but there is a limit to how much information we can fit into each pop-up, really with just space for the witness's written/spoken testimony.

To demonstrate the sighting density in each state, I now want to generate a map of North America where each of the 50 United States are shaded according to how many Bigfoot encounters have occured within it. If I were lucky, there would be no Bigfoot sightings in Alaska and I could simply use the United States map provided by the "maps" R package;

```{r}
bigfoot %>%
  filter(state == "Alaska" | state == "Hawaii") %>%
  group_by(state) %>%
  summarize(
    n_sightings = n()
  )
```

But alas. We would be missing out on a substantial 20 sightings by excluding Alaska.

So instead, I will need to use the R package USAboundaries to get the geometries for each of the 48 contiguous states and also Alaska and join this data with a dataset containing the number of sightings per state.

```{r}
states <- states(cb = TRUE)

library(USAboundaries)

states <- us_states() %>%
  filter(name != "Hawaii", name != "Puerto Rico") %>%
    st_crop(xmin = 130, xmax = -60,
            ymin = 0, ymax = 70)

sightings_by_state <- bigfoot %>%
  group_by(state) %>%
  summarize(
    n_sightings = n()
  )

sightings_joined <- full_join(states, sightings_by_state, join_by(name == state))

ggplot(data = sightings_joined) +
  geom_sf(aes(fill = n_sightings)) +
  scale_fill_viridis_c(direction = -1) +
  coord_sf(xlim = c(-170, -60), ylim = c(20, 73)) +
  labs(fill = "Number of Sightings") +
  theme_void()
```

Now this is where it really got interesting. Washington appears to have *by far* the most Bigfoot sightings, clocking in at over 600. What makes Washington different from other states? How dramatic is its lead? For my next visualization, I want to investigate these questions by generating a line graph that depicts the rise of bigfoot sightings over time in the top five states with the most sightings over all. We conveniently already have a date variable in the original dataset, so it's really just a matter of figuring out how to generate a line plot that shows the **cumulative** number of cases at each "point."

```{r}
sightings_by_state %>%
  slice_max(order_by = n_sightings, n = 5)
```

```{r}
sightings_over_time <- bigfoot %>%
  filter(!is.na(date)) %>%
  filter(state == "Washington"|
         state == "California" |
         state == "Ohio" |
         state == "Florida" |
         state == "Oregon") %>%
  mutate(
    date = ymd(date)
  ) %>%
  group_by(state, date) %>%
  summarize(
    n_sightings = n()
  ) %>%
  mutate(
    cumulative_sightings = cumsum(n_sightings)
    )


ggplot(sightings_over_time, aes(x = date, 
                                y = cumulative_sightings, 
                                color = state)) +
  geom_line(linewidth = 1) +
  labs(
    title = "Bigfoot Sightings Over Time",
    x = "Date",
    y = "Total Sightings",
    color = "State"
  ) + 
  scale_color_manual(values = c("red","seagreen2","slateblue2","thistle2","yellow")) +
  scale_x_date(limit=c(as.Date("1950-01-01"),as.Date("2026-04-24"))) +
  labs(caption = "*7 sightings have been excluded as they occured before 01-01-1950. Axis scaled for ease of reading.") +
  theme_minimal() 
```

With this visualization, we can make a few minor observations. The first is that Washington has, more or less, always led the 50 states in Bigfoot sightings. The only time where Washington was not in the lead was in the 60s, when California briefly had more. Other interesting observations; ever since approximately 2005, sightings in both Oregon and California have slowed to an almost pleateaued rate. This is especially interesting because of these states' close proximity to Washington, which is still experiencing sightings at a higher rate than the other 4 states.

Regardless of the reasoning behind this pattern, with the large sample size to choose from, I believe Washington is a good candidate for my final visualization; a linear regression seeking to predict the "quality" of a sighting. We have been given 3 levels of quality; Class A (best), Class B (fair), and Class C(poor). For the sake of simplicity, however, I am going to code the quality as 1 (Class A, best), 2 (Class B, fair), and 3(Class C, worst).

Because the BFRO cites visibility as a major determinant of an encounter's quality, I am going to primarily use variables that I think impact a person's ability to accurately see a sasquatch through any kind of weather phenomenon, such as **visibility, precip_intensity,** and **precip_type.** Additionally, I want to use plotly to add some hovertext just giving some context to the points on plot, such as which **county** the encounter occurred in,

First, I have to tidy the data by filtering out just the cases in Washington, coding the **classification** variable as numeric, and coding NAs in the precip_type variable as "None," with the assumption (and hope) that most of them are coded as NA because there was no precipitation, not because the precipitation type was not recorded.

```{r}
bigfoot_quality <- bigfoot %>%
  filter(state == "Washington") %>%
  mutate(
    quality = case_when(
      classification == "Class A" ~ 1,
      classification ==  "Class B"~ 2,
      classification == "Class C" ~ 3
  ),
    precip_type = factor(case_when(
      precip_type == "rain" ~ "rain",
      precip_type == "snow" ~ "snow",
      is.na(precip_type) ~ "none"
    ))
  ) 

mod_quality <- lm(quality ~ visibility + precip_intensity + precip_type + precip_type:precip_intensity, data = bigfoot_quality)

pred_quality <- ggpredict(mod_quality, terms = c("visibility [fivenum]",
                              "precip_intensity [fivenum]",
                              "precip_type")) %>%
  as.data.frame(terms_to_colnames = TRUE) %>%
  as_tibble()

plot_quality <- ggplot(data = pred_quality, aes(x=visibility, y=predicted))+
  geom_line(aes(color = precip_intensity), linewidth = 1.5) +
  facet_wrap(~precip_type) +
  scale_color_brewer(palette = "Pastel1") +
  theme_dark() +
  geom_point(data = bigfoot_quality, aes(y=quality, 
                                         color = precip_type,
                                         Title = `title`,
                                         County = county,
                                         Classification = classification
                                         ), alpha = 0.5)

ggplotly(plot_quality, tooltip = c("Title", "County", "Classification"))
```

So this graph is a little difficult to interpret, given the nature of the data, *however*, it is possible to make some interesting observations. For example, we see that there is only one Class C encounter recorded in Washington, and it was on a clear day with 10/10 visibility. As for the lines; unsurprisingly, clear days were closer to a classification of Class A than rainy days, across all visibilities and precipitation intensities. All around, it does look like as visibility increases, more encounters tend towards Class A. Interestingly, snowy days tended to be more credible than rainy days. My guess is that some of the evidence that the BFRO considered came in the form of traces left in the snow , like large footprints (as referenced by 2/7 of the points in the "snow" panel). There is also far less vegetation in the winter to obscure the big hairy form of a Bigfoot through the branches.

## **Conclusions**

Overall, the main takeaways from this project were that Washington is the Bigfoot capital of the United States, having had the lead since 1970 and not giving up its title once in the 50+ years since. In the same visualization, we also noted that Oregon and California have both experienced a significant plateau since 2010, though the significance and cause of this pattern is unknown. Additionally, in a linear regression based on various climate variables, we also noticed that snow weather is significantly more likely to be considered credible than a rainy encounter, and that credibility goes *up* as the intensity of the snow increases.


